Stimjim: open source hardware for precise electrical stimulation

### 2019-07-01 benchmarking/paper/draft2.Rnw
added opa197 datasheet, updated analysis

In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015;siegle_neural_2015}. Examples in life sciences include software and hardware for:
\begin{itemize*}
\item recording or stimulating neurons (e.g. Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013})
\item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017})
\item quantifying chemical reactions or microbial growth in 96-well plates\cite{szymula_open-source_2019}
\item growing and evolving bacteria\cite{takahashi_low_2015} 
\item ecological monitoring (e.g., Audiomoth\cite{hill_audiomoth:_2018})
\end{itemize*}
We now add Stimjim to this growing body of open hardware. Stimjim replaces commercial \textit{in vivo} neural stimulators at a fraction of the cost, with improved programmability. Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.

@@

\begin{table*}[h]
\small{
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                                &               &                  & STG400x                           & PHM-15x                  & ISO-01B       \\
                                & Stimjim       & PulsePal 2       & (Multichannel Systems)            & (Med Associates)         & (NPI)         \\
\hline
 Output channels                & 2             & 4                &  2-8                              & 2                        & 1             \\ 
 Stimulus generator included    & Y             & Y                &  Y                                & Y                        & N             \\  
 Current output mode            & Y             & N                &  Y                                & Y                        & Y                        \\
 Voltage output mode            & Y             & Y                &  Y                                & N                        & Y                        \\
 Outputs electrically isolated  & Y             & N                &  Y                                & Y                        & Y                        \\
 Onboard measurement            & Y             & N                &  N                                & Optional                 & Y*             \\
 Compliance voltage             & $\pm$15~V (voltage) & $\pm$10~V  & $\pm$120~V                        & $\pm$45~V                & $\pm$100~V     \\
                                & $\pm$13.7~V (current)    &       &                                   &                          &                \\
 Fastest pulse                  & $\approx$30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
 Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
                                &               & \$695 (assembled,& \$7462  (4 channels)              &  for software)           &                \\
                                &               & from Sanworks)   & \$10327 (8 channels)              &                          &                \\
 Open source                    & Y             & Y                &  N                                &  N                       & N              \\
 \hline
\end{tabular}
}
\caption{\label{tab:stimulatorComparison} \textbf{Comparison of Stimjim vs other related hardware.} Cost was obtained as either the complete cost of the bill of materials (Stimjim and PulsePal) or quoted prices from distributors (STG400x, PHM-15x, ISO-01B). *ISO-01B has a compliance warning sound if the load resistance is too high, but does not report actual current measurements.} 
\end{table*}

<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,warning=FALSE,message=FALSE,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).  (B) Median IPI in a one-second pulse train measured on a high-speed data acquisition device, plotted against target IPI. Red dots show current output, black dots show voltage output. Solid line shows equality.  (C) Boxplot of IPI errors measured in a 1-second pulse train. Note that the due to the 2~MHz sampling rate of our test equipment, the box representing the interquartile range often has zero height. Whiskers are drawn to most extreme values, irrespective of interquartile range. (D) Example pulse trains with varying pulse width (PW). (E) Median measured PW vs target PW. Red dots show current output, black dots show voltage output. Solid line shows equality. (F) Boxplots of pulse width errors over 100 pulses. Conventions are the same as for (C). (G) Example pulses with varying amplitude. (H) Error in amplitude of voltage pulses (100~$\\mu$s, 1~kHz) vs target amplitude. (I) same as (H) but for current pulses. (J) Example of a fast biphasic (30~$\\mu$s/phase) pulse. Red line is a 500~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 5~V pulse. (K) Example of a small-amplitude pulse. Red line is a 20~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 0.2~V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=

@@

padj=-6
layout(matrix(c(rep(1:9, each=2),rep(10,3), rep(11,3)),4,6,byrow=TRUE), widths=c(0.8,0.8,1.2,0.8,1,1))
par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01,cex.axis=0.9)
matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
        main = "Varying frequency\n(Inter-pulse interval, IPI)"); 
mtext("A",side=2, padj=padj-0.2,las=2, line=0)
par(mar=c(4,3,2,1))
plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, yaxt='n',
  log='xy', main='', cex=0.75, xlab='Target IPI (ms)', ylab='Measured IPI (ms)')
axis(2,las=2)
points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e3, 
  col='red', cex=0.75)
abline(a=0,b=1)
mtext("B",side=2, padj=padj,las=2)
legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')

@@

gplots::plotCI(x,y[2,], li=y[1,], ui=y[3,], yaxt='n', lty=1, gap=0, sfrac=0,
  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, ylim=c(-4,4))
axis(2,las=2)
mtext("C",side=2, padj=padj,las=2)

@@

par(mar=c(4,3,2,1))
plot(paramMat[31:59,7]/1e3, sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e3/2,  yaxt='n',
  log='xy', main='', xlab='Target PW (ms)', ylab='Measured PW (us)', cex=0.75)
points(paramMat[31:59,7]/1e3, sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e3/2,
  col='red', cex=0.75)
abline(a=0,b=1)
axis(2,las=2)
mtext("E",side=2, padj=padj,las=2)
legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')

x = rep(paramMat[31:60,7]/1e3, each=2)
y = sapply(seq(61,120), function(i) quantile(res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7], c(0,0.5,1)))
gplots::plotCI(x,y[1,], li=y[1,], ui=y[3,], yaxt='n', range=0, lty=1, gap=0, sfrac=0,
  log='x', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2)
axis(2,las=2)
mtext("F",side=2, padj=padj,las=2)

@@

\textbf{Electrode implantation and ICSS}. Monopolar electrodes (Plastics1, #MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}. The ground was implanted in the contralateral cortex. Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation. Typical resistance between connector pins after implantation was 20-30~k$\Omega$. All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).

\textbf{Field stimulation experiments}. Platinum-iridium electrodes (Alpha Omega, #387-102S01-11, 250~$\mu$m diameter, Parylene C and Polyamide coated, 0.1~M$\Omega$) were placed.

@@

<<sliceExpt, eval=FALSE>>=
if (!require(readABF)){
  install.packages("devtools")
  devtools::install_github("Zabolekar/abfReader")
}

sliceDat = readABF::readABF("./slice2/27_06_19_0016.abf")$data[[1]][,1]

@@

plot(sliceDat[1:5e5], type='l')
abline(v=i[1:20], col='red')

m = sapply(i, function(j) sliceDat[j+(-50:500)])
image(m, zlim=c(-100,50))

matplot(1:nrow(m)/10-5, dendRites::offset(m,10), type='l', lty=1, col=1, xlab='time (ms)', ylab='Membrane potential (mV)', yaxt='n',bty='n')

aps = sapply(i, function(j) {
  x = sliceDat[j+50:500]
  length(getLocalMaxima(x,threshold=0))
})
aps = colSums(matrix(aps,nrow=5))
@

>>>>

In contrast, there has been a recent push within the scientific community to produce open labware - open source hardware and software replacements for a variety of common laboratory tasks\cite{pearce_open-source_2013,baden_open_2015;siegle_neural_2015}. Examples in life sciences include software and hardware for:
\begin{itemize*}
\item recording or stimulating neurons (e.g. Open Ephys\cite{siegle_open_2017,buccino_open_2018}, Cyclops\cite{newman_optogenetic_2015}, Neurorighter\cite{newman_closed-loop_2013})
\item amplifying DNA (e.g., OpenPCR\cite{noauthor_openpcr_nodate}) or preparing single cells for DNA sequencing \cite{booeshaghi_design_2019}
\item imaging microscopic samples (e.g., FlyPi\cite{chagas_100_2017}, Octopi\cite{li_octopi:_2019})
\item quantifying chemical reactions or microbial growth in 96-well plates\cite{szymula_open-source_2019}
\item growing and evolving bacteria\cite{takahashi_low_2015} 
\item ecological monitoring (e.g., Audiomoth\cite{hill_audiomoth:_2018})
\end{itemize*}
We now add Stimjim to this growing body of open hardware. Stimjim replaces commercial \textit{in vivo} neural stimulators at a fraction of the cost, with improved programmability. Furthermore, due to its entirely open design and software, Stimjim can be modified by end users to fit their specific needs.

@@

\begin{table*}[h]
\small{
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                                &               &                  & STG400x                           & PHM-15x                  & ISO-01B       \\
                                & Stimjim       & PulsePal 2       & (Multichannel Systems)            & (Med Associates)         & (NPI)         \\
\hline
 Output channels                & 2             & 4                &  2-8                              & 2                        & 1             \\ 
 Stimulus generator included    & Y             & Y                &  Y                                & Y                        & N             \\  
 Current output mode            & Y             & N                &  Y                                & Y                        & Y                        \\
 Voltage output mode            & Y             & Y                &  Y                                & N                        & Y                        \\
 Outputs electrically isolated  & Y             & N                &  Y                                & Y                        & Y                        \\
 Onboard measurement            & Y             & N                &  N                                & Optional                 & Y*             \\
 Compliance voltage             & $\pm$15~V (voltage) & $\pm$10~V  & $\pm$120~V                        & $\pm$45~V                & $\pm$100~V     \\
                                & $\pm$13.7~V (current)    &       &                                   &                          &                \\
 Fastest pulse                  & $\sim$30~$\mu$s      & 100~$\mu$s       & 20~$\mu$s                         & 60~$\mu$s               & 10~$\mu$s      \\
 Cost (USD)                     & \$194 (parts) & \$264 (parts)    & \$4131 (2 channels)               &  \$6211 (+ \$4438     & \$1708         \\
                                &               & \$695 (assembled,& \$7462  (4 channels)              &  for software)           &                \\
                                &               & from Sanworks)   & \$10327 (8 channels)              &                          &                \\
 Open source                    & Y             & Y                &  N                                &  N                       & N              \\
 \hline
\end{tabular}
}
\caption{\label{tab:stimulatorComparison} \textbf{Comparison of Stimjim vs other related hardware.} Cost was obtained as either the complete cost of the bill of materials (Stimjim and PulsePal) or quoted prices from distributors (STG400x, PHM-15x, ISO-01B). *ISO-01B has a compliance warning sound if the load resistance is too high, but does not report actual current measurements.} 
\end{table*}

<<benchmarkFigure, echo=FALSE,fig.height=8,fig.width=5,warning=FALSE,message=FALSE,fig.cap="\\label{fig:benchmarking} \\textbf{Electrical benchmarks show Stimjim provides microsecond temporal precision and millivolt- and microampere-scale amplitude precision.}  (A) Example pulse trains with varying inter-pulse interval (IPI).  (B) Median IPI in a one-second pulse train measured on a high-speed data acquisition device, plotted against target IPI. Red dots show current output, black dots show voltage output. Solid line shows equality.  (C) Plot of IPI errors measured in a 1-second pulse train. Points show median error, error bars indicate worst-case errors. (D) Example pulse trains with varying pulse width (PW). (E) Median measured PW vs target PW. Red dots show current output, black dots show voltage output. Solid line shows equality. (F) Plots of pulse width errors over 100 pulses. Conventions are the same as for (C). (G) Example pulses with varying amplitude. (H) Error in amplitude of voltage pulses (100~$\\mu$s, 1~kHz) vs target amplitude. (I) same as (H) but for current pulses. (J) Example of a fast biphasic (30~$\\mu$s/phase) pulse. Red line is a 500~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 5~V pulse. (K) Example of a small-amplitude pulse. Red line is a 20~$\\mu$A current pulse (with a 9.75 k$\\Omega$ resistive load), black line is a 0.2~V pulse.", fig.align='center', out.width='1\\linewidth', fig.env="figure">>=

@@

padj=-6
layout(matrix(c(rep(1:9, each=2),rep(10,3), rep(11,3)),4,6,byrow=TRUE), widths=c(0.8,0.8,1.2,0.8,1,1))
par(mar=c(4,1,2,1), mgp=c(2,0.3,0), cex.main=1.05, tck=-0.01,cex.axis=0.9)
matplot(1e3/fs*1:nrow(ex1), dendRites::offset(ex1,15), col=1, type='l', lty=1, bty='n', yaxt='n', ylab='', xlab='time (ms)',
        main = "Varying frequency\n(Inter-pulse interval, IPI)"); 
mtext("A",side=2, padj=padj-0.2,las=2, line=0)
par(mar=c(4,3,2,1))
plot(paramMat[1:30,3]/1e3, sapply(res[seq(1,60,by=2)], function(x) x$IPI.summary[3])*1e3, xaxt='n',yaxt='n',
  log='xy', main='', cex=0.75, xlab='Target IPI (ms)', ylab='Measured IPI (ms)')
axis(2,las=2)
axis(1, las=2)
points(paramMat[1:30,3]/1e3, sapply(res[seq(2,60,by=2)], function(x) x$IPI.summary[3])*1e3, 
  col='red', cex=0.75)
abline(a=0,b=1)
mtext("B",side=2, padj=padj,las=2)
legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')

@@

gplots::plotCI(x,y[2,], li=y[1,], ui=y[3,], lty=1, gap=0, sfrac=0,  xaxt='n',yaxt='n',
  log='x', main='', cex=0.75, xlab='Target IPI (ms)', ylab='IPI error (us)', col=1:2, ylim=c(-4,4))
axis(1,las=2)
axis(2,las=2)
mtext("C",side=2, padj=padj,las=2)

@@

par(mar=c(4,3,2,1))
plot(paramMat[31:59,7]/1e3, sapply(res[seq(61,118,by=2)], function(x) x$PW.summary[3])*1e3/2, xaxt='n',yaxt='n',
  log='xy', main='', xlab='Target PW (ms)', ylab='Measured PW (us)', cex=0.75)
points(paramMat[31:59,7]/1e3, sapply(res[seq(62,118,by=2)], function(x) x$PW.summary[3])*1e3/2,
  col='red', cex=0.75)
abline(a=0,b=1)
axis(1,las=2)
axis(2,las=2)
mtext("E",side=2, padj=padj,las=2)
legend('topleft', col=1:2, pch=1, legend=c("voltage","current"), bty='n')

x = rep(paramMat[31:60,7]/1e3, each=2)
y = sapply(seq(61,120), function(i) quantile(res[[i]]$PW*1e6/2 - paramMat[(i+1)/2,7], c(0,0.5,1)))
gplots::plotCI(x,y[1,], li=y[1,], ui=y[3,], lty=1, gap=0, sfrac=0,  xaxt='n',yaxt='n',
  log='x', main='', cex=0.75, xlab='Target PW (ms)', ylab='PW error (us)', col=1:2)
axis(1,las=2)
axis(2,las=2)
mtext("F",side=2, padj=padj,las=2)

@@

\textbf{Electrode implantation and ICSS}. Monopolar electrodes (Plastics1, \#MS303/2-AIU/SPC, coated stainless steel, 200~$\mu$m diameter) were implanted above the medial forebrain bundle according to the protocol in reference \cite{carlezon_intracranial_2007}. The ground was implanted in the contralateral cortex. Additionally, a 3D-printed headpost was affixed to the animal's skull by dental cement to enable head fixation. Typical resistance between connector pins after implantation was 20-30~k$\Omega$. All animal procedures were in accordance with guidelines established by the NIH on the care and use of animals in research and were confirmed by the Technion Institutional Animal Care and Use Committee (IL-012-01-18, valid until 10/4/2022).

\textbf{Field stimulation experiments}. Platinum-iridium electrodes (Alpha Omega, \#387-102S01-11, 250~$\mu$m diameter, Parylene C and Polyamide coated, 0.1~M$\Omega$) were placed.

@@

<<sliceExpt, eval=TRUE,echo=FALSE,message=FALSE>>=
if (!require(readABF)){
  install.packages("devtools")
  devtools::install_github("Zabolekar/abfReader")
}

sliceDat = readABF::readABF(paste(projectDirectory,"slice2/27_06_19_0016.abf",sep=''))$data[[1]][,1]

@@

#plot(sliceDat[1:5e5], type='l')
#abline(v=i[1:20], col='red')

m = sapply(i, function(j) sliceDat[j+(-50:500)])
image(x=1:551/10, z=m, zlim=c(-100,50),xlab='time (ms)')

#matplot(1:nrow(m)/10-5, dendRites::offset(m,10), type='l', lty=1, col=1, xlab='time (ms)', ylab='Membrane potential (mV)', yaxt='n',bty='n')

#aps = sapply(i, function(j) {
#  x = sliceDat[j+50:500]
#  length(dendRites::getLocalMaxima(x,threshold=0))
#})
#aps = colSums(matrix(aps,nrow=5))
@
